

\chapter{Data Representation}


\section{Introducing Binary}



Here we introduce some non-standard notation. 
We use these \textit{Oxford Brackets} 
to show we are assigning a meaning to the binary digits. 
The \textit{natural} (unsigned) assignment is denoted by subscript $N$. 
Suppose we have a binary digits \textsf{x} and the unsigned we associate is $x$.
We write this as: $\llbracket \text{\textsf{x}} \rrbracket_{n} = x$
where the $n$ denotes a \textit{natural} interpretation of the binary. 

\highlightdef{\textbf{Number Table Series Expansion}: 
$\llbracket \text{\textsf{x}} \rrbracket_{n} = \sum^{T-1}_{t=0} 2^{t} \textsf{x}_t$}

This is the most natural way to assign an 
arithmetical value to this sequence of digits. 

\frmrule

Proof of uniqueness.

\frmrule



\section{Excess-n}

We write this as: $\llbracket \text{\textsf{x}} \rrbracket_{ex(n)} = x$

The definition of excess-n works backwards from the definition 
of $\llbracket \text{\textsf{x}} \rrbracket_{n}$. 
If we compare $\llbracket \text{\textsf{x}} \rrbracket_{n}$ 
to $\llbracket \text{\textsf{x}} \rrbracket_{ex(n)}$,
then $\llbracket \text{\textsf{x}} \rrbracket_{n}$, from the point of view 
of our new representation, is \textit{in excess} of $n$ 
(has an extra of $n$). Written as an equation, the definition is as follows.

\highlightdef{\textbf{Excess-n}: 
$\llbracket \text{\textsf{x}} \rrbracket_{n}  = 
\llbracket \text{\textsf{x}} \rrbracket_{ex(n)} + n$}

The value, $x$, that we want excess-n to give for binary \textsf{x} 
is $n$ too much when we use the natural representation. It's \textit{in excess}. 
And so, to find $\llbracket \text{\textsf{x}} \rrbracket_{ex(n)}$, 
we simply subtract $n$ from the natural representation. 

\frmrule 

\begin{example}
Find $\llbracket \text{\textsf{1011}} \rrbracket_{ex(8)}$.
\end{example}

\frameans{}{3}

The normal interpretation \textsf{x} has an excess of 8 (having an extra of 8) 
in comparison with the excess-8 interpretation. 
So if we subtract 8 from $\llbracket \text{\textsf{x}} \rrbracket_{n}$, 
we will find the $\llbracket \text{\textsf{x}} \rrbracket_{ex(8)}$


\section{One's Complement}


\section{Two's Complement}

We write this as: $\llbracket \text{\textsf{x}} \rrbracket_{2c} = x$




% TODO: fix render
\highlightdef{\textbf{Two's Complement}: 
$ \llbracket \text{\textsf{x}} \rrbracket_{2c} = \left\{ 
  \begin{array}{ll}
    \llbracket \text{\textsf{x}} \rrbracket_{n} & 
    \quad \text{for} \quad 0 \leqslant \llbracket \text{\textsf{x}} \rrbracket_{n} < 2^{n-1}  \\
    \llbracket \text{\textsf{x}} \rrbracket_{n} - 2^{n} & 
    \quad \text{for} \quad 2^{n-1} \leqslant \llbracket \text{\textsf{x}} \rrbracket_{n} < 2^{n}
  \end{array} \right.$
}

\highlightdef{\textbf{Two's Complement Operation}: 
$\llbracket 2C(\text{\textsf{x}}) \rrbracket_{2c}  = 
- \llbracket \text{\textsf{x}} \rrbracket_{2c}$
}



\section{Ten's Complement}

We write this as: $\llbracket \text{\textsf{x}} \rrbracket_{10c} = x$


\section{Hexidecimal}


\section{General Base-n}

\highlightdef{\textbf{Number Table Series Expansion}: 
$\llbracket \text{\textsf{x}} \rrbracket_{r} = \sum^{N-1}_{t=0} r^{t} \textsf{x}_t$}