
\chapter{Division}




\section{Introducing Division in Rings}

What do we know about division from group theory?
When we add addition and subtraction into the mix, things become a bit more confusing. 
Division in rings is centred around the equation $ra = rb$. In particular, we want to know 
when we can cancel $r$ and when we cannot. 

It turns out if we have $r$, then it falls into certain exactly \textit{four} disjoint cases. 
In three cases, we can cancel $r$ but in one case we cannot. The rest of our work in this chapter 
will be looking at these four cases and showing that they are disjoint.

\begin{itemize}	
\renewcommand{\labelitemi}{$\Box$}
\item \textbf{Invertible.} In the first case, $r$ is invertible which means that we can 
cancel $r$ in $ra = rb$ to get $a = b$. This is obvious because we can left-multiply 
both sides of $ra = rb$ by $r^{-1}$ to get $a =b$. This first case we have seen from group theory. We didn't 
need any addition/subtraction to see that $a$ and $b$ must be equal.
\item \textbf{Non-invertible Cancellable.} In the second case, $r$ does not have a multiplicative inverse 
but can still be cancelled. It's pretty counter-intuitive seeing these elements and there aren't many 
rings that have these types of elements in them. There is no name given to these elements. 
So for a lack of a better name, we call them \textit{non-invertible cancellable elements}.
\item \textbf{Non-cancellable.} In the third case, we have \textit{Non-invertible Non-cancellable elements} 
or simple \textit{Non-cancellable}. This the bad case where we cannot cancel $r$ from $ra = rb$. 
\end{itemize}




\begin{figure}[h]
\begin{tikzpicture} [
	every node/.style={node_sty},
	edge from parent/.style={edge_sty},
	edge from parent path={(\tikzparentnode.south) -- ++(0, -0.2cm) -| (\tikzchildnode.north)},
	level/.style = {level distance = 0.6cm},
	level 1/.style={sibling distance=6cm},
	level 2/.style={sibling distance=3cm}
]
	\node (root) {$ra = rb$}
    child{
    	node (can) {$r$ invertible}
		child {node(inv) [text width=6em] {$r$ cancellable}}
	}
	child{
		node (ncan) {$r$ non-invertible}
		child {node (nz) [text width=6em] {$r$ cancellable}}
		child {node (z) [text width=8em] {$r$ non-cancellable}}
	};
	
\end{tikzpicture}
\end{figure}


\begin{example}
Here is an example of case 1. In $(\mathbb{R}, +, \times)$ we have the equation $5x = 5y$.
We can simply multiply both sides by the multiplicative inverse of 5 which is $\frac{1}{5}$, this gives us $x=y$. 
\end{example}

\begin{example}
Here is an example of case 2. Suppose in $(\mathbb{Z}, +, \times)$ we have the equation $2x = 2y$. This equation tells us that 
doubling the integer $x$ is the same as doubling the integer $y$. Intuitively we can tell that 
$x = y$. In this ring, it wouldn't make sense to double two \textit{different} integers and get the \textit{same result}. 
$2$ is not invertible in $(\mathbb{Z}, + \times)$ yet we know we should be able to cancel it from $2x = 2y$ to get $x = y$.
\end{example}


So cancelling things is not as simple as multiplying by inverse elements. The previous example 
demonstrated that we can cancel elements that are not invertible. This is part of the reason why 
division in rings is more complex than division in groups. 

\begin{example}
Case 3 is the most interesting. Rearranging gives us $r(a-b) = 0_R$. The next step
you would \textit{expect} would be to say that either $r = 0$, or $a-b = 0$. 
This step is called \textit{the zero-factor property}.
However, in general, doing this step here is wrong. It turns out 
that \textit{sometimes} $r(a-b) = 0_R$ without either $a$ or $a-b$ being zero at all! 
For example look at $(\mathbb{Z}_6, +_6, \times_{6})$.
We have $3 \times (4-2) = 0$. Yet $r = 3$ is not zero and $4-2$ is not zero. 
So going from $3 \times (4-2) = 0$ to saying $r = 0$ or $a - b = 0$ is wrong. 
\end{example}

The zero factor property is $r(a-b) = 0_R$ implies $r = 0_R$ or $a-b = 0_R$
This can be rewritten in logic as not($r(a-b) = 0_R$ and not ($r = 0_R$ or $a-b = 0_R$))

Now in the non-cancellable case, if we assume the zero-factor property, we will get a contradition.
And so we will see that zero-factor property is not true. 
Hence  $r(a-b) = 0_R$ and not ($r = 0_R$ or $a-b = 0_R$)
in other words $r(a-b) = 0_R$ and $r \neq 0_R$ and $a-b \neq 0_R$.

This is the condition that tells us that we cannot cancel $r$.

\highlightdef{\textbf{Non-cancellable Property}:  $r(a-b) = 0_R$ and $r \neq 0_R$ and $a-b \neq 0_R$}
The non-cancellable property tells us when we cannot cancel $r$ from $r(a-b) = 0_R$ when we 
it is because $r$ and $a-b$ multiply to give zero (and neither of them are 0). 






\section{Zero Divisors}




\highlightdef{$r \neq 0_R$ is a \textbf{Zero-divisor} iff $\exists a \exists b$ st $a-b \neq 0_R$ and $r(a-b) = 0_R$}
In other words, $r$ is called a \textit{zero-divisor} is there is at least one case 
where it cannot be cancelled in $r(a-b) = 0_R$. This is when $r$ times $a-b$ gives zero but neither $r$ nor 
$a-b \neq 0_R$ is zero. Suppose we let $s = a - b$, then we can rewrite the definition as follows: 

\highlightdef{$r \neq 0_R$ is a \textbf{Zero-divisor} iff $\exists s$ st $s \neq 0_R$ and $rs = 0_R$}

This form is equivalent but a simpler to work with for later proofs. However it no longer captures the 
idea that we are trying to cancel $r$ from the equation $ra = rb$.

\begin{example}
Show that 4 is a zero-divisor in 
\end{example}




It turns out that being invertible and being a zero-divisor are two disjoint cases.
 The diagram below shows this. This makes things simpler. Let's prove this result. 


\begin{figure}[h]
\begin{tikzpicture} [
	every node/.style={node_sty},
	edge from parent/.style={edge_sty},
	edge from parent path={(\tikzparentnode.south) -- ++(0, -0.2cm) -| (\tikzchildnode.north)},
	level/.style = {level distance = 0.6cm},
	level 1/.style={sibling distance=6cm},
	level 2/.style={sibling distance=3cm}
]
	\node (root) {$ra = rb$}
    child{
    	node (can) {$r$ invertible}
		child {node(inv) [text width=6em] {$r$ cancellable}}
	}
	child{
		node (ncan) {$r$ non-invertible}
		child {node (nz) [text width=6em] {$r$ cancellable}}
		child {node (z) [text width=8em] {$r$ non-cancellable}}
	};
	
\end{tikzpicture}
\end{figure}



\highlightdef{$r$ cannot be both a \textit{unit} and a \textit{zero-divisor}}

Assume that $r$ is both a \textit{unit} and a \textit{zero-divisor}. 
By definition of zero-divisor we have $\exists a \exists b$ st $a-b \neq 0_R$ and $r(a-b) = 0_R$. 
But we assumed that $r$ is a unit and so $r^{-1}$ exists. Using this, we can multiply $r(a-b) = 0_R$ on 
both sides to get $a-b = 0_R$. This contradicts our assumption that $a-b \neq 0_R$. 
Hence $r$ cannot be both a \textit{unit} and a \textit{zero-divisor}.

So being a zero divisor of some kind and a unit at the same time isn't possible. 
These events are mutually exclusive. You either one, or the other, or neither. 
THe three cases of cancellation vs non-cancellation of $r$ correspond exactly with whether 
$r$ is a zero-divisor or a unit or neither. This is shown in the table below.

Recall that $(x \boxtimes)$ is using Haskell's \textit{section notation} and denotes 
the function $f(y) = x \boxtimes y$. Similarly, $(\boxtimes x)$ denotes $f(y) = y \boxtimes x$. To check 
that an element is a left zero-divisor, 






\section{Non-commutative Ring Division}

\begin{example}
Support xyz = 0. Explain why x is a right zero divisor.
\end{example}

Notice how we didn't need to add brackets to xyz. Multiplication in rings is always associative.
So x(yz) is the same as (xy)z. 



\begin{example}
Give a formal definition of a \textit{left proper zero divisor}. 
\end{example}


If our multication isn't commutative then we need to be careful as to whether we use $f_x$ or $g_x$. 
However if multication is commutative, then every left zd is also going to be a right zd.
And every right zd will also be a left zd. So saying left/right is redundant.
So, for zds in \textit{commutative rings}, we say that they are simply \textbf{a zero divisor}.  





\highlightdef{$r$ cannot be both a \textit{left-unit} and a \textit{left zero-divisor}}

Assume $r$ is a \textit{left-unit} and a \textit{left zero-divisor}.
By definition of a left zero-divisor we have $\exists a \exists b$ st $a-b \neq 0_R$ and $r(a-b) = 0_R$. 
But $r$ is also a left-unit so has a \textit{left-inverse} $r_L$ exists. 
Using this, we can multiply $r(a-b) = 0_R$ on the left to give $(a-b) = 0_R$
This contradicts our assumption that $a-b \neq 0_R$. 
Hence $r$ cannot be both a \textit{unit} and a \textit{zero-divisor}.

\highlightdef{$r$ cannot be both a \textit{right-unit} and a \textit{right zero-divisor}}

This intuitively follows by symmetry. This can be proved is the same way the left case was proved.





%\highlightdef{\textbf{$\mathbb{Z}_d$ unit condition}: x is a unit iff $\text{hcf}(x,d) = 1$ \\
%\textbf{$\mathbb{Z}_d$ zd condition}: x is a zd iff $\text{hcf}(x,d) \neq 1$  }




\section{Fields and Domains}

\highlightdef{A \textbf{Domain} is a ring with no proper zero-divisors}

\highlightdef{Every \textit{domain} satisfies the \textit{zero-product property}}

This is saying that when we have $ab = 0$ then neither $a$ nor $b$ are left/right zero-divisors.
It is not the case that $a \neq 0$ and can sent

\highlightdef{A \textbf{Commutative Domain} is a commutative ring with no proper zero-divisors}

For historical reasons, Commutative Domains are better known as \textit{Integral Domains}. 
Why Integral? This is likely because the ring of integers is an domain. 
When we multiply numbers together,we don't get zero - unless we are multiplying by zero. 
So saying integral is a reminder that domains are a bit like $\mathbb{Z}$




\highlightdef{An \textbf{Field} is a ring with no proper zero-divisors and no proper non-units}

An integral domain cuts out one area. A field cuts out two areas.
In a domain, you won't get any zd's but you may get some non-units.

$F = F^{*} \cup {0}$
$F = F^{*} \cup F_{\text{pnu}} \cup  \left\{0\right\}$


\begin{example}
Find what's wrong with the following diagram for a field.
\end{example}
\frameans{You may assume $R \neq {0}$}{Zero is inside zero-divisors.}


